---
title: "Calculation of winter variability: X days below Y °C"
author: "Jens Joschinski"
date: "February 1, 2018"
output: 
    md_document:
        variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache=TRUE)
#setwd(dirname(dirname(dirname(getwd()))))
library(RCurl)
library(readr)
library(data.table)
library(textreadr)
library(tidyr)
library(dplyr)
library(stringr)
library(magrittr)
library(imputeTS)
threshold<-50
t2<-10
```
# General description  
## Aim  
The aim of this project is to correlate climate variability with variability in seasonal timing. Is the slope in seasonal responses a bet-hedging trait, i.e., is it adaptive to spread one's timing in more variable conditions?  


### Overview  
The script uses the daily minimum and maximum temeratures for all months, years and climate stations all over the world from the GHCN dataste (~12million X 31 days).These were pre-processed with PERL and saved as two txt files. The two datasets are now opened, cleaned, and the mean of min and max is calculated for each day. The resulting workspace with one matrix of 11milx31days is saved.  
Further processing uses smaller chunks of data at a time (30 chunks of 1k climate stations). These are subsequently read in and transformed to differen format (so that 1 row has temperature of a whole year). Then I calculate winter onset as xth day that temperature falls below y°C. For each station I calculate mean winter onset (weighted by available data per year), standard deviation in winter onset, and winter predictability: I make a linear regression through the last 31 days before winter onset of each year, and calculate the slope. The inverse of the standard deviation of this slope estimate across years is winter predictability.Finally, all data of one station is put into a single vector, and I apply a non-linear least square regression: The data is expected to follow a sine-curve with a period of 365 days, and I record for each station amplitude (difference between summer and winter), phase angle (day of peak summer) and intercept (mean temperature). The results (mean winter onset, sd, correlation, nls-parameters) are saved in a text file. All of this is done on all chunks of data, and the results are appended in a single file.



### Specific description  

The data was generated with R version `r getRversion()`. It uses the GHCN-daily dataset by NOAA:
```{r, echo =F}
text<- read_document("01climate_data/01raw/ghcnd_all/ghcnd-version.txt")
print (text[1])
```


### Script  


#### Read climate station data and clean up 

Reading data and removing stations for which only mintemp or only maxtemp exists
```{r read_n_clean}
converted<-fread("01climate_data/02processing/001data_conversion/converted_max.txt", sep ="|", na.strings=c("NA","-9999"),verbose=F)

length(unique(converted$V1))
conv_table<-table(converted$V1)
#absolute minimum shuold be 3 years of data with at least half of the year covered
#this definition is not complete (e.g. does not exclude 1 day per month for 20 years) but reduces the dataset a bit 
inlist<-conv_table[conv_table>=18] 
converted<-converted[converted$V1 %in% names(inlist),]
                                                                                                                                                                                                                                                                                                    

converted2<-fread("01climate_data/02processing/001data_conversion/converted_min.txt", sep ="|", na.strings=c("NA","-9999"),verbose=F)
conv_table2 <- table(converted2$V1)
inlist2<-conv_table2[conv_table2>=18]
converted2<-converted2[converted2$V1 %in% names(inlist2),]

#now in both files stations with <3 years are removed: 2753 from max temps, 2633 from min temps

converted2<-converted2[converted2$V1 %in% names(inlist),] 
#min temps consists now only of stations that are also in max temps
converted<-converted[converted$V1 %in% names(inlist2)]
#and max temps only has stations that are also in min temps
length(unique( converted$V1))
length(unique(converted2$V1)) #same
```

An older version of this script removed climate stations below 20°N or 18°N here. But some study sites are at ~ 18°N, and the interpolation to climate stations requires a 5° radius. to be sure, I now try at 0°N

```{r remove_under20}
url<-"ftp://ftp.ncdc.noaa.gov/pub/data/ghcn/daily/ghcnd-stations.txt"
#this dataset is fixed-width delimited, requiring a few additional steps
locations<-read.fwf(
  file=url
  ,sep="!",na.strings=c("NA","-999.9"), #sep = ! because ! does not exist in dataset - > dataset is fixed-width and should have no additional separators
  widths=c(11, 9, 10, 7,2,35)
)

badlist <- locations[locations[,2]<0,1]  #latitudes smaller 0°N
rm(locations)
rm(url)
converted<-converted[!(converted$V1 %in% badlist),]
#removed 1,163,477 lines = 3,445 stations
converted2<-converted2[!(converted2$V1 %in% badlist),]
#removed 1,185,911 lines = 3,445 stations
rm(inlist)
rm(inlist2)
rm(conv_table)
rm(conv_table2)
rm(badlist)
rm(text)
```


removing any month that occurs in only one of the two datasets (nrow should be the same for both)

```{r remove_uniquemonths}
converted<-unite_(converted,"new_id",c("V1","V2","V3"),remove=FALSE)
converted2<-unite_(converted2,"new_id",c("V1","V2","V3"),remove=FALSE)


converted2<-converted2[converted2$new_id %in% converted$new_id]
converted<-converted[converted$new_id %in% converted2$new_id]
nrow(converted)
#we are now down to 11.7 million rows in each dataset

converted<-arrange(converted, new_id)
converted2<-arrange(converted2, new_id)


#calculate average Temperature:
#(max+min)/2
join<-converted
join[,5:ncol(join)]<-(converted[,5:ncol(converted)]+converted2[,5:ncol(converted)])/2
converted<-join
rm(join)
rm(converted2)
converted<-converted[,-1]
names(converted)<-c("ID","year","month",1:31)
mymonths <- c("07Jan","08Feb","09Mar",
              "10Apr","11May","12Jun",
              "01Jul","02Aug","03Sep",
              "04Oct","05Nov","06Dec")
converted$month<-mymonths[converted$month]  #when it is sorted alphabetically it now starts in july, and ends in jun
rm(mymonths)

#some more cleaning
navector<-rowSums(!is.na(converted[,4:ncol(converted)]))
converted<-converted[navector>0,]
conv_table <- table(converted$ID)
inlist<-conv_table[conv_table>=18] 
converted<-converted[converted$ID %in% names(inlist),]
#11,715,375 lines, 28963 stations

save.image(file = "converted.RData") 
#row 1         ID year month   1     2     3   4     5     6   7   8   9    10  11  12  13  14  15
#1 ACW00011604 1949 07Jan 253 258.5 252.5 258 255.5 255.5 253 242 247 230.5 228 236 214 225 228
#     16  17    18  19    20    21    22  23  24    25  26    27    28  29  30    31
#1 244.5 239 241.5 233 244.5 247.5 244.5 239 239 233.5 250 244.5 241.5 247 239 244.5
```

#### data structure  
Due to problems with memory allocation, the big (26 million rows) dataset needs to be processed piece-wise. I chose 30 chunks of 1000 climate stations each. 
The following steps have to be done on that 1000 station dataset

* change data structure from 
    station 1 year 1 month 1 day 1,2,3..31
    station 1 year 1 month 2 day 1,2,3..31  
  to:  
    station 1 year 1 day 1,2,3.. 365
    station 1 year 2 day 1,2,3.. 365  
    
* make subset with 1 station and
  * calculate winter onset in each year  
  * calculate mean winter onset (weighted by number of days per year)  
  * calculate weighted standard deviation in winter onset  
  * calculate standard deviation in slopes of temperature decline before winter  
  * change structure to station 1 day 1....20,000 and make nls regression  

changing the data structure:
```{r transform_matrix}
turn_around <- function(station){
station<-
  station %>% 
  gather(key="day",value="temperature",-(1:3)) %>% #makes one very long dataframe with 2 columns
  mutate (day=str_pad(day,2,pad="0")) %>%  #this makes leading zeros where they are needed
  mutate (temperature=as.numeric(temperature)) %>%
  arrange (ID,year,month,day) %>%
  unite_("un",c("month","day")) %>%  #makes new col with new ID
  spread(key="un",value="temperature")#puts in new wide format with 365 columns (month-day)

#length(station$`08Feb_29`) #not many, due to leapyears
x<-NA
for(i in 1:ncol(station)){x[i]<-length(unique(station[,i]))}
station<-station[-which(x==1)]  #dates like 30 feb are erased
return(station)
}

```



### nls regression  (though that part wont be needed anymore)
The following chunk has the function to apply a nls regression on the climate data of each station. The daily temperatures over ~20 years are expected to follow a sine-curve pattern with a period of 1 year. The curve is determined by the following parameters:  
* a *constant c* that defines the average temperature throughout the year. It is around 10°C in temperate climates, around 25°C at the equator.  
* the *amplitude A*, which quantifies the difference between winter and summer temperatures. A should decrease with proximity to the poles.  
* the *phase angle phi*. Phi defines at what time of the year maximum temperatures occur. It should be close to midsummer in all stations of the northern hemisphere. june 25 is the julian day 176/177. This corresponds to a phase angle of 170/365 \*(2\*pi) = 2.92.

```{r nls_regression, include=F}
get_nls <- function (vals,s_A=400,s_phi=pi/2,s_c=200){#s_... are starting values for nls function
 if (is.na(length(vals))|| length(vals)<1000){return (NA)} else{ #when there is insufficient data, the function will return "NA"
  x<-1:length(vals)
  data<-data.frame(x,vals)

  #make nls regression for function 'y=Amplitude * cosine(period*x + phase angle) + intercept'
  res <- nls(vals ~ A*cos(x*2*pi/366+phi)+C, data=data, start=list(A=s_A,phi=s_phi,C=s_c),
            upper=c(500,2*pi,400),algorithm="port",lower=c(0,0,-400)) 
#366 days, non-leap-years have an NA at feb29

  return(coef(res))
  }
}


#wrapping function that allows continuing if errors occur in one of the stations
wrap_error = function (vals,s_A,s_phi,s_c) {
  tryCatch(get_nls(vals,s_A,s_phi,s_c),
                     warning = function (w)  {return(FALSE)},
                     error   = function (e)  {return(FALSE)})
}

```

### actual calculation  
Do the calculations, piecewise for 29 chunks of data; save the results
```{r}   

x<-unique(converted$ID)
for (loop in 1:29){ #this loop goes through the 29 chunks of 1k stations (last chunk has only 963 entries)
station<-converted[converted$ID %in% x[((loop-1)*1000+1):(1000*loop)],]
station<-turn_around(station)

#remove years with <182 days of data
n<-rowSums(!is.na(station))
station<-station[n>184,] #2 more becasue they always have id and year

#remove stations with less than 3 years
temp<-table(station$ID)
temp <-as.data.frame(t(temp)) 
temp<-temp[temp$Freq<3,2]
temp<-droplevels(temp)
station<-station[!(station$ID %in% temp),]
station<-droplevels(station)

#vectors to store results
mean_winter<-rep(NA,length(unique(station$ID)))
sd_winter<- rep(NA,length(unique(station$ID)))
nyears<-rep(NA,length(unique(station$ID)))
p<-rep(NA,length(unique(station$ID)))
beta_y<-rep(NA,length(unique(station$ID)))
beta_m<-rep(NA,length(unique(station$ID)))
#A<-rep(NA,length(unique(station$ID)))
#phi<-rep(NA,length(unique(station$ID)))
#int<-rep(NA,length(unique(station$ID)))


for (i in 1:length(unique(station$ID))){ #this loop goes through each of the 1 k stations
  
  ####read in station####
  id<-unique(station$ID)[i]
  s<-filter(station,ID==id)
  ndays<-rowSums(!is.na(s))-2
  #format of the data: year 1, day 1...365
  #                    year 2, day 1...365
  
   
  
  ####calculate winter arrival#####
 # make a binomial matrix with all Temps<threshold = 1, all others =0(including NA)

  bino<-rep(0,nrow(s)*(ncol(s)-2))
  bino[s[,3:368]<threshold]<-1
  bino<-matrix(data=bino,nrow=nrow(s),ncol=ncol(s)-2)

  #exclude years that never reach winter (less than x days below y°C)
  used<-rowSums(bino)>t2
  
  #exclude stations that reach winter in less than 3 years (this is different from having less than 3 ys of data)
  if (sum(used) <3) {
    mean_winter[i]<-NA
    sd_winter[i]<-NA
     next
  }
  
  bino<-bino[used,]
  s<-s[used,]
  w_on<-rep(NA,nrow(bino))
  ndays<-ndays[used]
  
  #calcualte row-wise cumulative sums
  #function cumsum works only on columns, so matrix needs transposing
  bino <- 
    bino %>%
    t %>% 
    as.data.frame %>%
    cumsum %>%
    t %>%
    as.data.frame
  
  bino<-bino==t2 #turns it into TRUE/FALSE
 w_on<- max.col(bino, ties.method="first")#first "true" ="threshold reached" in each year
 nyears[i] <- length(w_on)
  
  ##calculate mean and sd in winter arrival (both weighted)####
 
  mean_winter[i]<-weighted.mean(w_on,ndays)
  weights<-ndays/366
  upper <- sum(weights*(w_on-mean_winter[i])^2)
  lower <- sum(weights)-1
  sd_winter[i] <- sqrt(upper/lower)
  
  #clean up memory####
  rm(upper)
  rm(lower)
  rm(weights)
  rm(ndays)
  rm(bino)


  ##calculate standard deviation in slopes of last 30 days before winter onset####
  #idea: transpose and use gather to make 1 long vector ("day1:day 10,000" only that they are not called like that)
        #cut at (winter onset - 30) and winter onset of each year (multiple pieces)
        #clean incomplete data
        #make correlations of temperature ~ time to get the slope of each year
        #calculate sd of the slopes between years
  
  #transpose
  rownames(s)<-1:nrow(s) #these are years 1:x
  n<-gather(as.data.frame(t(s[3:368]))) #turns into long format with 2 columns:
  #key = 1,1,1...(366 times), 2,2,2...(366 times),..,40,40... (assuming there are 40 years)
  #value = temperature year 1, day 1:366, year 2, day 1.... 
  n$xvec<-rep(1:(ncol(s)-2),length(unique(n$key))) #xvec is days after 1st july, so it repeats 1:366, 40 times
  n$key<-as.integer(n$key)
  
  #cut
  n<-filter(n,xvec<w_on[key] & xvec >= w_on[key]-31)  #now dataset contains only last 31 days before winter onset in that respective year
  
  #clean
  n2<-group_by(n,key)
  sm<-summarise(n2,sum(!is.na(value))>10) #makes a tibble with 2 cols, key and usable==T or F
  usable<-pull(sm[sm[,2]==T,1]) #makes a vector with all years that have >10 days
  n<-n[n$key %in% usable,]

  #slope estimates
  slopes<-rep(NA,length(usable))
  for (keygroup in usable){ 
    y<-filter(n,key == keygroup)
    slopes[keygroup]<-coef(lm(y$value~y$xvec))[2]
  }
  #translation: for each year take data of that year, save slope estimate from lm
  p[i]<-sd(slopes,na.rm=T)


  
  #make non-linear least-square regression ####

# The function tries up to three times per climate station, with different starting parameters.
 n<-gather(as.data.frame(t(s[3:ncol(s)])))
notneeded2<-function(){
 parms<-wrap_error(n$value,s_A=400,s_phi=pi/2,s_c=200)
  if (parms[1]==FALSE){parms<-wrap_error(n$value,s_A=40,s_phi=pi,s_c=200)}
  if (parms[1]==FALSE){parms<-wrap_error(n$value,s_A=40,s_phi=2*pi,s_c=200)}
  if (parms[1]==FALSE){print (paste("failed in ",i));parms<-c(-9999,-9999,-9999)}

 A[i]<-parms[1]
 phi[i]<-parms[2]
 int[i]<-parms[3]
 
  }
  
 ####  spectral density estimate: beta ###
#two ways to calculate beta are tested:1. over the whole year, 2. over last 60 days before winter onset.

#whole year
 obs<-n[,2]
obs<-as.data.frame(obs)
ys           <- nrow(obs)/366

# fill in gaps by interpolation
obs          <-na.interpolation(obs)

# append day
obs           <- cbind(obs,rep(1:366,ys))
colnames(obs) <- c("measurement","day")

# extract daily mean from each observations  
daily.average <- ave(obs$measurement,obs$day)
obs$norm      <- obs$measurement - daily.average

# spectral density
dens          <- spectrum(obs$norm,plot=F,method="pgram")
# select the interval [1/n,.5/month] (i.e. ignore frequencies below the resolution of the data and above a 2-month period)
freq          <- dens$freq[dens$freq >= 1/(60*ys) & dens$freq<= 1/59]
spec          <-dens$spec[dens$freq >= 1/(60*ys) & dens$freq<= 1/59]

# linear regression
if(length(freq)<5){next
  }else{
reg <- lm(log10(spec)~log10(freq))
beta_y[i]<--coef(reg)[2]
}

#2. 2 months
  #idea: transpose and use gather to make 1 long vector ("day1:day 10,000" only that they are not called like that)
        #cut at (winter onset - 60) and winter onset of each year (multiple pieces)
        #clean incomplete data
        #calculate colour of noise on these chunks
  s<-s[w_on>67,]
  w_on<-w_on[w_on>67]
  s<-droplevels(s)
  if(length(w_on[w_on>60])==0){next}
  #transpose
  rownames(s)<-1:nrow(s) #these are years 1:x
  n<-gather(as.data.frame(t(s[3:368]))) #turns into long format with 2 columns:
  n<-droplevels(n)
  #key = 1,1,1...(366 times), 2,2,2...(366 times),..,40,40... (assuming there are 40 years)
  #value = temperature year 1, day 1:366, year 2, day 1.... 
  n$xvec<-rep(1:(ncol(s)-2),length(unique(n$key))) #xvec is days after 1st july, so it repeats 1:366, 40 times
  n$key<-as.integer(n$key)
  
  #cut
  n<-filter(n,xvec<w_on[key] & xvec >= w_on[key]-60)  #now dataset contains only last 60 days before winter onset in that respective year
  
  #clean
  n2<-group_by(n,key)
  sm<-summarise(n2,sum(!is.na(value))>40) #makes a tibble with 2 cols, key and usable==T or F
  usable<-pull(sm[sm[,2]==T,1]) #makes a vector with all years that have >10 days
  if(length(usable)<1){next}
  n<-n[n$key %in% usable,]

  obs<-n[,2]
obs<-as.data.frame(obs)
ys           <- nrow(obs)/60

# fill in gaps by interpolation
obs          <-na.interpolation(obs)

# append day
obs           <- cbind(obs,rep(1:60,ys))
colnames(obs) <- c("measurement","day")

# extract daily mean from each observations  
daily.average <- ave(obs$measurement,obs$day)
obs$norm      <- obs$measurement - daily.average

# spectral density
dens          <- spectrum(obs$norm,plot=F,method="pgram")
# select the interval [1/n,.5/month] (i.e. ignore frequencies below the resolution of the data and above a 2-month period)
freq          <- dens$freq[dens$freq >= 1/(60*ys) & dens$freq<= 1/59]
spec          <-dens$spec[dens$freq >= 1/(60*ys) & dens$freq<= 1/59]

if(length(freq)<5){next
  }else{
# linear regression
reg <- lm(log10(spec)~log10(freq))
beta_m[i]<--coef(reg)[2]
}

#abline(lm(log10(dens$spec)~log10(dens$freq)))



  } #next station


#####gather and save results####
reslist<-data.frame(unique(station$ID),mean_winter,sd_winter,nyears,beta_y,beta_m)

names(reslist)<-c("ID","meanwinter","sd winter","nyears","predictability_year","predictability_month" )
write.table(reslist,"results.txt",append=TRUE,quote=FALSE,row.names=FALSE,col.names=FALSE)
} #next chunk
```


In the next step the results have to be combined with data from empirical studies. this is in another script (folder 03Analysis)

```{r}
freqs<-n[,2]
x<-1:length(freqs)
freqs[x%%365==1]
ms<-rep(NA,365)
for ( i in 1:365){
  ms[i]<-mean(freqs[x%%366==i],na.rm=T)
}
plot(freqs)
points(ms,col=2)
freq2<-freqs
for ( i in 1:365){
  freq2[x%%365==i]<-freq2[x%%365==i]-ms[i]
}


p<-plot(spectrum(freq2[!is.na(freq2)]))
p$pred<-predict(lm(log(p$spec)~p$freq))
points(exp(p$pred)~p$freq)
beta<- exp(coef(lm(log(p$spec)~p$freq))[2])
```

