---
title: "Calculation of winter variability: X days below Y °C"
author: "Jens Joschinski"
date: "February 1, 2018"
output: 
    md_document:
        variant: markdown_github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache=TRUE)
#setwd(dirname(dirname(dirname(getwd()))))
library(RCurl)
library(readr)
library(data.table)
library(textreadr)
library(tidyr)
library(dplyr)
library(stringr)
library(magrittr)
threshold<-50
t2<-10
#load("converted.RData")
```
# General description  
## Aim  
The aim of this project is to correlate climate variability with variability in seasonal timing. Is the slope in seasonal responses a bet-hedging trait, i.e., is it adaptive to spread one's timing in more variable conditions?  


### Overview  
The script opens a pre-processed (with PERL) dataset, which includes daily minimum and maximum temeratures for all months, years and stations all over the world (~65million X 31 days). It converts the dataset into [5 million x 365 days] format, and splits it in smaller chunks and saves these as new files.
Then all small files are read in subsequently to calculate winter onset in each year, mean and sd of winter onset, and correlation of winter onset to temperature 1,2 and 3 weeks before winter onset. The results are again stored in text files.
In the end, the results are read in again, and merged with climate station metadata (needs to connects with the NOAA server to get a list of available climate stations together with their latitudes and longitudes. This is the only time a connection to the server is required).



### Specific description  

The data was generated with R version `r getRversion()`. It uses the GHCN-daily dataset by NOAA:
```{r, echo =F}
text<- read_document("01raw/ghcnd_all/ghcnd-version.txt")
print (text[1])
```

* read all stations
* remove unneccessary stations (<3 ys, <20°N)
- convert to matrix [stations,1:365], june first
- cut into N pieces
- save as file

- for each file: 
- get winter onset for each year
- get sample size for each year
- get mean onset, sd
- save results: vector means,sds, n_years;

### Script  


#### Read climate station data and clean up 

Reading data and removing stations for which only mintemp or only maxtemp exists
```{r read_n_clean}
converted<-fread(paste(getwd(),"/02processing/001data_conversion/converted_max.txt",sep=""), sep ="|", na.strings=c("NA","-9999"),verbose=F)

length(unique(converted$V1))
conv_table<-table(converted$V1)
inlist<-conv_table[conv_table>=3]
converted<-converted[converted$V1 %in% names(inlist),]


converted2<-fread(paste(getwd(),"/02processing/001data_conversion/converted_min.txt",sep=""), sep ="|", na.strings=c("NA","-9999"),verbose=F)
conv_table2 <- table(converted2$V1)
inlist2<-conv_table2[conv_table2>=3]
converted2<-converted2[converted2$V1 %in% names(inlist2),]

#now in both files stations with <3 years are removed: 700 from max temps, 611 from min temps

converted2<-converted2[converted2$V1 %in% names(inlist),] 
#min temps consists now only of stations that are also in max temps
converted<-converted[converted$V1 %in% names(inlist2)]
#and max temps only has stations that are also in min temps
length(unique( converted$V1))
length(unique(converted2$V1)) #same
```

Removing climate stations below 20°N

```{r remove_under20}
url<-"ftp://ftp.ncdc.noaa.gov/pub/data/ghcn/daily/ghcnd-stations.txt"
#this dataset is fixed-width delimited, requiring a few additional steps
locations<-read.fwf(
  file=url
  ,sep="!",na.strings=c("NA","-999.9"), #sep = ! because ! does not exist in dataset - > dataset is fixed-width and should have no additional seps
  widths=c(11, 9, 10, 7,2,35)
)

badlist <- locations[locations[,2]<20,1]
rm(locations)
rm(url)
converted<-converted[!(converted$V1 %in% badlist),]
#removed 1,163,477 lines = 3,445 stations
converted2<-converted2[!(converted2$V1 %in% badlist),]
#removed 1,185,911 lines = 3,445 stations
rm(inlist)
rm(inlist2)
rm(conv_table)
rm(conv_table2)
rm(badlist)
rm(text)
```


removing any month that occurs in only one of the two datasets (nrow should be the same for both)

```{r remove_uniquemonths}
converted<-unite_(converted,"new_id",c("V1","V2","V3"),remove=FALSE)
converted2<-unite_(converted2,"new_id",c("V1","V2","V3"),remove=FALSE)


converted2<-converted2[converted2$new_id %in% converted$new_id]
converted<-converted[converted$new_id %in% converted2$new_id]
#we are now down to 11.4 million rows in each dataset

converted<-arrange(converted, new_id)
converted2<-arrange(converted2, new_id)
#(max-min)/2 + min

converted[,5:ncol(converted)]<-(converted[,5:ncol(converted)]-converted2[,5:ncol(converted2)])/2+converted2[,5:ncol(converted2)]
rm(converted2)
converted<-converted[,-1]
names(converted)<-c("ID","year","month",1:31)
mymonths <- c("07Jan","08Feb","09Mar",
              "10Apr","11May","12Jun",
              "01Jul","02Aug","03Sep",
              "04Oct","05Nov","06Dec")
converted$month<-mymonths[converted$month]  #when it is sorted alphabetically it now starts in july, and ends in jun
rm(mymonths)
save.image(file = "converted.RData")
```



```{r}
turn_around <- function(station){
station<-
  station %>% 
  gather(key="day",value="temperature",-(1:3)) %>% #makes one very long dataframe with 2 columns
  mutate (day=str_pad(day,2,pad="0")) %>%
  mutate (temperature=as.numeric(temperature)) %>%
  arrange (ID,year,month,day) %>%
  unite_("un",c("month","day")) %>%
  spread(key="un",value="temperature")#puts in new wide format with 365 columns (month-day)

#ncol(station)
#length(station$`08Feb_29`) #not many, due to leapyears
#data structure
#station[1:4,1:10]
x<-NA
for(i in 1:ncol(station)){x[i]<-length(unique(station[,i]))}
station<-station[-which(x==1)]  #dates like 30 feb are erased
return(station)
}

```

```{r}

for (loop in 1:29){
station<-converted[converted$ID %in% unique(converted$ID)[(loop-1)*1000+1:1000*loop],]
station<-turn_around(station)


mean_winter<-rep(NA,length(unique(station$ID)))
sd_winter<- rep(NA,length(unique(station$ID)))
cor_7<-rep(NA,length(unique(station$ID)))
cor_14<-rep(NA,length(unique(station$ID)))
cor_21<-rep(NA,length(unique(station$ID)))
nyears<-rep(NA,length(unique(station$ID)))
c<-rep(NA,length(unique(station$ID)))
for (i in 1:length(unique(station$ID))){
  #make a subset with 1 station
  #remove years with <182 days of data
  #skip station with <3 years, and years where winter is not reached
  #for the rest make a binomial matrix with all Temps<threshold = 1, all others =0(including NA)
  #calculate winter onset: point in each year where cumulative sum is reached
  #calculate weighted mean winter (weighted by no of days/year) and sd (also weighted)
  #calculate correlation of winter onset and temp 1,2,3 weeks ago
  
  id<-unique(station$ID)[i]
  s<-filter(station,ID==id)
  
  ndays<-rowSums(!is.na(s))-2 #number of days with data per year, -2 because need to substract id and year
  s<-s[ndays>182,3:ncol(s)]
  ndays<-ndays[ndays>182]
  
  if (nrow(s)<3) {
    mean_winter[i]<-NA
    sd_winter[i]<-NA
    next  #skip stations with less than 3 years of +-complete data
  }
  
  
  ##calculate winter arrival
  bino<-s
  bino[bino<threshold]<-1
  bino[bino>=threshold]<-0
  bino[is.na(bino)]<-0
  rownames(bino)<-1:nrow(bino)
  
  used<-rowSums(bino)>t2#exclude years that never reach winter (less than x days below y°C)
  if (sum(used) <3) {
    mean_winter[i]<-NA
    sd_winter[i]<-NA
     next
  }
  bino<-bino[used,]
  w_on<-rep(NA,nrow(bino))
   
  #for each row calculate date when cumsum = t2
   for (j in 1:nrow(bino)){
     w_on[j]<-min(which(cumsum(as.integer(bino[j,]))==t2))
   }

  ##calculate mean and sd in witner arrival
  ndays<-ndays[used]
  mean_winter[i]<-weighted.mean(w_on,ndays)
  weights<-ndays/366
  upper <- sum(weights*(w_on-mean_winter[i])^2)
  lower <- sum(weights)-1
  sd_winter[i] <- sqrt(upper/lower)
  rm(upper)
  rm(lower)
  rm(weights)
  rm(ndays)
  rm(bino)

  #wcalculate correlation with x days ago
  m<-matrix(data=c(which(used==T),w_on),nrow=length(w_on),ncol=2)
  m2<-matrix(data=c(which(used==T),w_on-7),nrow=length(w_on),ncol=2)
  m2[m2<1]<-NA
  m3<-matrix(data=c(which(used==T),w_on-14),nrow=length(w_on),ncol=2)
  m3[m3<1]<-NA
  m4<-matrix(data=c(which(used==T),w_on-21),nrow=length(w_on),ncol=2)
  m4[m4<1]<-NA
  
  w<-s[m]
  w_7<-s[m2]
  w_14<-s[m3]
  w_21<-s[m4]

  cor_7[i]<-cor(w[!is.na(w_7)],w_7[!is.na(w_7)])
  cor_14[i]<-cor(w[!is.na(w_14)],w_14[!is.na(w_14)])
  cor_21[i]<-cor(w[!is.na(w_21)],w_21[!is.na(w_21)])
  nyears[i] <- length(w_on)
  
  ##calculate pearsons R along last 30 days before winter onset
  rownames(s)<-1:nrow(s)
  n<-gather(as.data.frame(t(s))) #turns into long format
  n$xvec<-rep(1:ncol(s),length(unique(n$key))) #xvec is days after 1.july
  n$key<-as.integer(n$key)
  n<-filter(n,xvec<w_on[key] & xvec >= w_on[key]-31)  #now dataset contains only last 30 days before winter onset in that respective year
  n2<-group_by(n,key)
  n<-filter(n2,sum(!is.na(value))>10)  #removes years that have less than 10 days of data in the 30 days before winter onset

  #plot(n$value~n$xvec,bg=n$key,pch=21)
 c[i]<- cor(n$value,n$xvec,use="na.or.complete")
}


reslist<-data.frame(unique(station$ID),mean_winter,sd_winter,cor_7,cor_14,cor_21,nyears,c)

names(reslist)<-c("ID","meanwinter","sd winter", "cor 1wk","cor2wk","cor3wk","nyears","cor")
write.table(reslist,"results.txt",append=TRUE,quote=FALSE,row.names=FALSE,col.names=FALSE)
}
```


```{r}
reslist<-read.table("results.txt")
r2<-merge(locations,reslist,by=1)
r2$cor<-r2$cor+1
r2<-r2[!is.na(r2$cor),]

r2<-r2[r2$V2<60,]
plot(r2$V2~r2$V3,bg=rgb(r2$cor,r2$cor,0,maxColorValue = 1.612),pch=22,col=NA,cex=0.4)
plot(r2$cor~r2$meanwinter)#cor becomes weaker in warmer climates
plot(r2$cor~r2$V4)#does not decrease with altitude
plot(r2$cor~r2$V2)#decreases with proximity equator
plot(r2$cor~r2$`sd winter`) #not strongly correlated to sd winter
cor(r2$cor,r2$`sd winter`)
plot(r2$cor~r2$nyears)
```

dataset is now sorted jun - jun, matrix of form
station1 y1 1:365
station1 y2 1:365
for each station
  make subset
  make wvector with sample sizesy per year sum (!is.na(row))
  get binomial matrix
  calculate winter based on binomial matrix
  erase binomial matrix
  ->vector with winter onsets
  
 1 take onset-vector, n-vector to calculate weighted mean and weighted sd (2 values)
2 make matrix (winter onset, winter onset - 7,14,21 days)
  calculate cor (onset, other cols) (3 values)
  3 melt to get station 1:10000 days, do nls to get AMP,PHI,C (3 values)
  save values in matrix, 
  go on with next station
  
  
  
##Quality control  

The following part will append all data from one station to a single vector and make a non-linear least square regression. The data is expected to follow a sine curve with a period of ~1 year.


####Function for creating single vector  

The function "daily_t" will use the data provided by one station. It will append all daily temperature recordings of all years (up to 145 years \* 12 months \* 31 days ) into a single vector, filling up with NAs as needed. 


```{r vectorize}
daily_t <- function (station){
  station<-station[arrange(station[,2],station[,3],na.last=F),] #sorts by year and month
  station<-droplevels(station)
  if (length(unique(station$year))<3){vals<- (NA)} #quality control
  else {
    vals<-NA
    #for each year:
    for(y in 1:length(unique(station$year))){
      year<-station[station[,2]==unique(station[,2])[y],] #reduces dataset to 1 station, 1 year          (max 31*12 points)
      
      #make 12*31 matrix and fill it with daily data
      mat <- matrix(NA,12,31)
  
      for (i in 1:nrow(year)){ #i=month
        mon<-as.numeric(year[i,3])
        mat[mon,1:31]<-as.numeric(year[i,4:34])
      }
    vals<-c(vals,as.numeric(t(mat)))
    }
  return (vals)
  }
}


```



####Function for nonlinear least-square regression  

The function 'get_nls' takes the daily temperatures of one station (which is supplied as single vector), and applies a non-linear least squares model which estimates intercept, phase angle and amplitude of a sine curve.

```{r nls_regression}
get_nls <- function (vals,s_A=400,s_phi=pi/2,s_c=200){#s_... are starting values for nls function
  if (length(vals)==1){return (NA)} else{ #when there is insufficient data, the function daily_t will return "NA", so vals gets length 1. Otherwise vals will have length of at least 2*370(many of which may be NA)
  x<-1:length(vals)
  data<-data.frame(x,vals)

  #make nls regression for function 'y=Amplitude * cosine(period*x + phase angle) + intercept'
  res <- nls(vals ~ A*cos(x*2*pi/372+phi)+C, data=data, start=list(A=s_A,phi=s_phi,C=s_c),
            upper=c(500,2*pi,400),algorithm="port",lower=c(0,0,-400)) 
  #372 and not 365.25, because a vector of 12*31 was used before 
  #(filling up e.g. 31.feb with NA). This made coping with leap years easier

  co <- coef(res)
  return(co)
  }
}
```



####applying the functions on the dataset

The following chunk will apply a nls regression on the climate data of each station. The daily temperatures over ~20 years are expected to follow a sine-curve pattern with a period of 1 year. The curve is determined by the following parameters:  
* a *constant c* that defines the average temperature throughout the year. It is around 20°C in temperate climates, around 30°C at the equator.  
* the *amplitude A*, which quantifies the difference between winter and summer temperatures. A should decrease with proximity to the poles.  
* the *phase angle phi*. Phi defines at what time of the year maximum temperatures occur. It should be close to midsummer in all stations of the northern hemisphere, and around midwinter in the southern hemisphere. june 25 is the julian day 176/177. BEcause the function daily_t assumes 12*31 = 372 days, it is around day 170 in this dataset. This corresponds to a phase angle of 170/372 \*(2\*pi) = 2.87.  The phase angle of midwinter is accordingly, 368/372\*(2\*pi)=6.22.

The function tries up to three times per climate station, with different starting parameters.

```{r quality-control}
#,error=TRUE}
phi=NA
A=NA
C=NA
lat=NA
lon=NA
reg<-data.frame(lon,lat,A,phi,C)

wrap_error = function (vals,s_A,s_phi,s_c) {
  tryCatch(get_nls(vals,s_A,s_phi,s_c),
                     warning = function (w)  {return(FALSE)},
                     error   = function (e)  {return(FALSE)})
}

for (i in 1:length(unique(testset$ID))){ #= for each climate station

temp_set<-testset[testset$ID==unique(testset$ID)[i],]
vals<-daily_t(temp_set)
if (length(unique(vals))>100){ #insufficient number of years in functon daily_t will make length (vals)=1; but it is still possible that there are 3+ years filled with NA (less than 100 data points)

  parms<-wrap_error(vals,s_A=400,s_phi=pi/2,s_c=200)
  if (parms[1]==FALSE){parms<-wrap_error(vals,s_A=40,s_phi=pi,s_c=200)}#print ("retry1");
  if (parms[1]==FALSE){parms<-wrap_error(vals,s_A=40,s_phi=2*pi,s_c=200)}#print ("retry2");
  if (parms[1]==FALSE){print (paste("failed in ",i));parms<-c(-9999,-9999,-9999)}

} else  {
  parms<-rep(NA,5)
}
reg[i,1]<-temp_set[1,35] #longitude
reg[i,2]<-temp_set[1,36]# latitude
reg[i,3]<-parms[1] # Amplitude
reg[i,4]<-parms[2] #phase angle
reg[i,5]<-parms[3]#intercept
reg[i,6]<-temp_set[1,1]#ID
}
reg<-reg[is.na(reg[,3])==F,]
```


####Plotting the results:

1. average Temperature
```{r T_avg}
restore <- reg
reg$C[reg$C<(-200)]<-(-200) #caps some very low temperatures, makes plotting gradient nicer

plot(reg[,1]~reg[,2],bg=rgb(reg[,5]+abs(min(reg[,5])),1,reg[,5]+abs(min(reg[,5])),maxColorValue = max(reg[,5])+abs(min(reg[,5]))),pch=21,col=NA,cex=0.5,
     main = "Average temperature(C)",xlab="°E",ylab="°N")
for (i in seq(min(reg[,5]),max(reg[,5]),1)){
  points(x=-150 + i/10,y = 0, bg= rgb(i+abs(min(reg[,5])),1,i+abs(min(reg[,5])),maxColorValue = max(reg[,5])+abs(min(reg[,5]))),col=NA,pch=22)
}
text(-160,-5,min(reg$C)/10)
text(-110,-5,max(reg$C)/10)
#hist(reg[,5],breaks=100)
reg<-restore
#mean Temperature behaves as expected
```


2. temperature amplitude
```{r t_amplitude}
restore <- reg
reg$A[reg$A>150]<-150 #caps some very high amplitudes
plot(reg[,1]~reg[,2],bg=rgb(reg[,3],1,reg[,3],maxColorValue = max(reg[,3])),pch=21,col=NA,cex=0.5,
     main = "Amplitude(A)",xlab="°E",ylab="°N")
for (i in seq(0,max(reg[,3]),1)){
  points(x=-150 + i/5,y = 0,bg=rgb(i,1,i,maxColorValue = max(reg[,3])),col=NA,pch=22)
}#left: A = 0, right: A = 24°
text(-150,-5,0)
text(-120,-5,max(reg[,3]/10))

#hist(reg[,3])

reg<-restore
#amplitude is 0 at equator, as expected.
#amplitude increases northwards to >15° diff in summer-winter
#In south, there are only data points up to 20°S in this sample dataset, but the trend is the same
```


3. phase angle
```{r phase_angle}
plot(reg[,1]~reg[,2],bg=rgb(reg[,4],1,reg[,4],maxColorValue = 2*pi),pch=21,col=NA,cex=0.5,
        main = "Phase angle",xlab="°E",ylab="°N")
for (i in seq(0,2*pi,0.1)){
  points(x=-150 + i*5,y = 0,bg=rgb(i,1,i,maxColorValue = 2*pi),col=NA,pch=22)
}#left: phi = 0, right: phi=2*pi
#hist(reg[,4])
#Due to the circular nature of the dataset, phase angles of phi=0 and phi=2*pi are equal. 
#better representation

plot(reg[,1]~reg[,2],bg=rgb(cos(reg[,4])+1,1,cos(reg[,4])+1,maxColorValue = 2),pch=21,col=NA,
       cex=0.5, main = "Phase angle",xlab="°E",ylab="°N")
for (i in seq(0,2*pi,0.1)){
  points(x=-150 + i*5,y = 0,bg=rgb(cos(i)+1,1,cos(i)+1,maxColorValue = 2),col=NA,pch=22)
}

#hist(cos(reg[,4]))

#phase angles behave as expected. mean of phi in the northern hemisphere:
mean(reg[,4][reg[,1]>20]) #this is shortly before midsummer
```


According to the subset, the data behaves just as expected.   
* amplitude decreases towards equator  
* amplitude is lower (less seasonality) at coastlines (e.g. Australia, west coast USA)  
* mean temperature increases towards equator  
* mean temperature drops in mountain ranges  
* phase angle is close to midsummer in northern latitudes, and -180° in southern latitudes



The remaining part of the script will be done on northern hemisphere only (>20°N), as there is no empirical data for southern hemisphere.


##calculate winter variability

The location metadata has been erased in the meantime, because it takes quite a bit of memory. Reload it:

```{r get_coordinates_again,include=FALSE}
url<-"ftp://ftp.ncdc.noaa.gov/pub/data/ghcn/daily/ghcnd-stations.txt"
#this dataset is fixed-width delimited, requiring a few additional steps
locations<-read.fwf(
  file=url
  ,sep="!",na.strings=c("NA","-999.9"), #sep = ! because ! does not exist in dataset - > dataset is fixed-width and should have no additional seps
  widths=c(11, 9, 10, 7,2,35)
)
names(locations) <- c("ID","lat","lon","alt","name","?")
```


A bit of cleaning...


```{r cleanup,echo=FALSE}
old<-testset #the full dataset, which includes equator and southern hemisphere; not needed anymore
newset<-old[old$lat>20,]
rm(testset)
gc()
testset<-newset[newset$ID==unique(newset$ID)[10],]
#now newset is the full dataset that is needed for analysis, testset a small subset for testing purposes
sfile<-unique(newset$ID)
```



####General idea

It is inconvenient that the year ends around midwinter. To calculate winter arrival, it makes more sense to have one "year" going from june to june:

```{r plot_idea}
plot(daily_t(testset),type="l")
#mean phase angle is at 2.78. 2.78/(2*pi)*372 =165 => highest temperature at day 165
points(x=165+372*(1:100),y= daily_t(testset)[165+372*(1:100)],col=2) #okay, dataset can be cut in this positiont
```

####How this is done

* process data from one station to get a single temperature vector (there is already a function for that, called daily_t)
* get rid of the first half year (which lies before the first red dot), and of the last half year (this does not contain full 12 months of information) 
* reassemble the temperature vector into a matrix, with each row representing 12 months (june to june) of data.
* For each year(row), count the number of days which have data
* calculate winter onset for each year:
   + make a binomial matrix, with all T<y = 1, all other T=0
   + calculate cumulative sums of each row
   + note the day where cum_sum > X
* calculate standard deviation and mean winter onset, weighted by number of days with data



####Function to cut and reassemble into binomial matrix

This function takes the looong temperature vector of one station (daily temperature of several years), cuts the head and tail (half a year each), puts it back into matrix format (1 row = 1 year), and transforms it into a binomial matrix


```{r get_binomial_matrix}
reassemble_cd <- function (station,degrees=50){ #method to calculate winter onset based on accumulation of days cooler than 50centidegree (5°C)
  #expected input: a vector of length 372*n. Should be of the form: for n years{temperatures from (Jan to Dec)}. 372, because the function daily_t uses 12*31 days for calculation (filling up with NAs)
  
  #error handling
  if (length(station[is.na(station)==F])<100){
    return (NA) #error if <100 days in total
  } else if ((length(station)-1)%%372 !=0 ){
    print("ERROR")
    return(NA) #error if there are incomplete years
  } else if (length(station)<=754){
    print ("ERROR")
    return(NA)#error if there are <3 years
  } else{
    
    #cutting and reassembling
    station <- station[2:length(station)] #function daily_t() has appended an NA to each year in the beggining
    nyears <- length(station)/372 #should be an integer
    station<-station[165:(length(station)-208)] #temperatures from first winter to summer are irrelevant, calculation begins with first summer; the last year ends in december, but data would be needed until june-->last year cannot be used for calculations
    
    
    #conversion to binomial
    station[station<degrees]<-1
    station[station>1]<-0
    m<-matrix(station,(nyears-1),372,byrow=T)
    return(m) #output: a matrix with 372 cols, n-1 rows(first half of first year, and 2nd half of last year removed)
  }
}

```



####Function to calculate number of days with data

The script will later calculate the mean winter onset at each station. Because data quality differs among years, the mean needs to be weighted by the number of days with data per year. To do that, one needs to actually have the number of days:


```{r, count_NAs}
row_na<-function (inp_matrix){
  navec<-NA #this means na_vector
  if (is.matrix(inp_matrix)==TRUE){
    navec <- rowSums(!is.na(inp_matrix))
   # for (i in 1:nrow(inp_matrix)){
   #   navec[i]<-length(inp_matrix[i,][is.na(inp_matrix[i,]==TRUE)])
   # }
  }
  return(navec)
}
```


####Function to test whether the cumulative sum of occurences of y reaches x = 10

```{r cumulator}
calc_cumul <- function (inp_matrix,thres=10, conv_NA =TRUE){ #takes a matrix as input, calculates rowwise cumulative sums, and returns the position on which a threshold was first reached in each row 
  
  #input: any 2-dimensional matrix, a threshold
  #output: sum_vector
  
  #error handling
  if (is.matrix(inp_matrix)==FALSE){stop("no matrix provided in function calc_cumul")}
  if (conv_NA ==TRUE){
  vector<-is.na(inp_matrix)
  inp_matrix[vector]<-0
  }
  if (NA %in% inp_matrix){ #This part will only run if conv_NA ==F and there are nevertheless NA in the matrix
      stop("NA found in function calc_cumul. Try conv_NA=TRUE")
  }
  
  sum_vector <- rep(NA,nrow(inp_matrix))
  for (m_row in 1:nrow(inp_matrix)){
    temporary <-cumsum(inp_matrix[m_row,])
    if  (length(temporary[temporary==thres])>0){#only continue if threshold is reached
      sum_vector[m_row]<-min(which(temporary==thres))  #Cumsum increases 'from left to right' in the matrix. But if there are zeros in the original matrix, the cumsum may stay on one value, so threshold will occur several times in a row. the min() makes sure only the first value is taken 
    } #else{sum_vector stays NA}
  }
  return (sum_vector)
}
```


####Function that puts all these functions for one station together

```{r winter_one_station}
do_all<-function(station,x=10,y=50){
  temp_set<-newset[newset$ID==unique(newset$ID)[station],] #read all data from one station
  
  vals<-daily_t(temp_set)#convert data frame into single vector: (day 1:372)*years
 #this returns: either a matrix, or NA
  
  n_t<-reassemble_cd(vals,y)# convert into binomial matrix of form {1:372}*(years-1)
  #this returns either a matrix or NA, input NA => output NA
  
  ndays<-row_na(n_t) #this vector describes how many temperature measurements are available per year for this station
  
  if ((is.null(nrow(n_t))==F) && (nrow(n_t)>1)){#calculation for this station only if functions "temp_set" and "reassemble" did not return {NA}
    out<- calc_cumul(n_t,x) #vector describing on which day y was reached for the xth time
  #error if no matrix as input, or if NA provided and setting conv_NA ==F
  }else{
    out<-NA
  }
  return(list(out,ndays))
}

```



####Function to calculate weighted means

The built-in function cannot handle NA

```{r weighted_means}
weighted_means<-function(resm,nm){ #calculate row-wise means of a matrix, weighted by another matrix, and ignoring any NAs
  ms<-rep(NA, nrow(resm))
  for (i in 1:nrow(resm)){
    worked<-which(!is.na(resm[i,1:ncol(resm)]))
    if (length(worked)>3){
      ms[i]<-weighted.mean(resm[i,worked[worked>3]],nm[i,worked[worked>3]])
    }
  }
  return(ms)
}
``` 

####function to calculate standard deviation
```{r sd}
standard_dev<-function(resm){ #calculate row-wise sd of a matrix, ignoring any NAs
  ms<-rep(NA, nrow(resm))
  for (i in 1:nrow(resm)){
    worked<-which(!is.na(resm[i,1:ncol(resm)]))
    if (length(worked)>3){
      ms[i]<-sd(resm[i,worked[worked>3]])
    }
  }
  return(ms)
}
``` 


####wrapping function to run the code for all stations


```{r winter_all_Stations}
loop_stations<-function(x=10,y=50){
#create a matrix:  9000 rows (1 for each station), 140 columns (1 for each year)+3 columns for ID,latitude, longitude. Each cell has the winter onset in one year
resmat<-matrix(NA,length(unique(newset$ID)),max(newset$year)-min(newset$year)+3)
nmat  <-matrix(NA,length(unique(newset$ID)),max(newset$year)-min(newset$year)+3)
lat_lon<-locations[locations$ID %in% sfile,1:3]
#table(sfile==lat_lon[,1])#always true
resmat<-as.data.frame(resmat)
nmat<-as.data.frame(nmat)
resmat[,1]<-sfile
resmat[,2]<-lat_lon[,2]
resmat[,3]<-lat_lon[,3]
nmat[,1:3]<-resmat[,1:3]
#loop through all stations, calculate winter onset given x and y values
for (station in 1:length(sfile)){
  r<-do_all(station,x,y)
  r[[1]][r[[2]]<100]<-NA #remove all years with less than 100 days of data
  resmat[station,4:(length(r[[1]])+3)]<-r[[1]]
  nmat  [station,4:(length(r[[2]])+3)]<-r[[2]]
}

final<-resmat
final2<-nmat
return(list(final,final2))
}

```



####running the actual code

This chunk tries 4 different parameter combinations
```{r run_the_code}
final_10_50<-loop_stations(x=10,y=50)
final_5_20<-loop_stations(x=5,y=20)
final_20_100<-loop_stations(x=20,y=100)
final_5_0<-loop_stations(5,0)

```

*The results are stored in lists with 2 entries: final[[1]] is the matrix with winter onsets in each year (columns) and for each station (rows), final[[2]] is a matrix with the according sample size (number of days per year)*




####plotting results


The follwoing chunk plots mean winter onset (weighted by sample size within each year, i.e. days with data), then sd(winter onset), and lastly the coefficient of variation, sd/mean. This is done for four parameter combinations. 


```{r plot_the_results}

lis<-list(final_10_50,final_5_20,final_20_100,final_5_0)
x<-c(10,5,20,5)
y<-c(50,20,100,0)
CV<-list(NA,NA,NA,NA) # will be used in later chunk


for (i in 1:length(lis)){
  input <- lis[[i]]
  subtitle<-paste(x[i],"days below ", y[i], "°C") #will be needed in figures
  
  input[[1]]$means<-weighted_means(input[[1]],input[[2]])
  input[[1]]$sds<-standard_dev(input[[1]])
  input<-input[[1]]
  input<-input[!is.na(input$means),]
  plot(input[,2]~input[,3],bg=rgb(0,input$means,0,maxColorValue=372),col=NA,main="weighted mean day of winter onset",sub=subtitle,pch=22,cex=0.8)
   
  input$sds[!is.finite(input$sds)]<-NA
  input<-input[!is.na(input$sds),]
  input<-input[input$sds>0,]
  plot(input[,2]~input[,3],bg=rgb(0,input$sds,0,maxColorValue=max(input$sds)),col=NA,main="Sd of winter onset",sub=subtitle,pch=22,cex=0.8)
  
  input$CV<-input$sds/input$means
  input$CV[!is.finite(input$CV)]<-NA
  input<-input[!is.na(input$CV),]
  CV[[i]]<-as.data.frame(input$CV,input$V1)
  plot(input[,2]~input[,3],bg=rgb(0,input$CV,0,maxColorValue=max(input$CV)),col=NA,main="Sd/mean of winter onset",sub=subtitle,pch=22,cex=0.8)
}
```

```{r mins}
minesweep_r <- function (input_vector){
  mines_vector <- rep(0,1000)
  mines_vector[sample(1000,10)]<-1
  if (sum(mines_vector[input_vector])>0){print ("you lost")} else {print ("you won")}
}
minesweep_r(input_vector=sample(1000,990))
```


The results of the 4 runs are very similar, and the sd and CV plots are very dark. this is because the standard deviation is rather long-tailed, caused by the northernmost stations. Cropping it a bit to get a better view (last run only):

```{r plot_cropped}
cropped<-input
cropped$sds[cropped$sds>50]<-50
cropped$CV[cropped$CV>0.5]<-0.5
  plot(cropped[,2]~cropped[,3],bg=rgb(0,cropped$sds,0,maxColorValue=max(cropped$sds)),col=NA,main="Sd of winter onset",sub=subtitle,pch=22,cex=0.6)
  plot(input[,2]~input[,3],bg=rgb(0,cropped$CV,0,maxColorValue=max(cropped$CV)),col=NA,main="Sd/mean of winter onset",sub=subtitle,pch=22,cex=0.6)
```



####comparing the 4 runs

```{r comparison}
m<-merge(CV[[1]],CV[[2]],by=0) #by=0 merges by rownames
m2<-merge(CV[[3]],CV[[4]],by=0)
m<-merge(m,m2,by=1) # 1, because they have now a separate column with names
names(m) =c("ID","10_50","5_20","20_100","5_0")
plot(m[,2:5])

```

These look quite similar, only the parameter run "20 days below 10°C" sticks out as different. But this parameter combination is really quite different from the others. 


####weighted standard deviation  

The number of days with data is likely different among years. How is the overall data quality?


```{r ndays}
cut<-final_10_50[[2]][,4:(ncol(final_10_50[[1]]))] #removes the columns mean and sd
mark<-!is.na(cut)
test<-cut[mark]
hist(test,breaks=100,xlab="Days",ylab ="Frequency",main ="Frequency distribution: Number of days/year")
length(test[test<300])/length(test)
# 18% <300


```


While most years have complete data, a significant proportion (>20%) is incomplete. the between-years standard deviation hence needs to be weighed by the reliability of each year (sample size within year).  

####calculation of weighted sd

```{r calc_sdweighed}
weighted<-list(NA,NA,NA,NA)
for (i in 1:length(lis)){
  input <- lis[[i]] #lis has the 4 runs stored
  weighted[[i]]<-data.frame(NA,NA,NA)
  names(weighted[[i]])<-c("ID","m","sd")
  
  for (j in 1:nrow(input[[1]])){
    mean_w_on<-NA #will temporarily store mean winter onset of this station
    wsd <- NA     #same for weighted sd
    
    w_on<-input[[1]][j,4:(ncol(input[[1]])-3)] #a vector with winter onsets of each year
    weights<-input[[2]][j,4:(ncol(input[[2]])-3)] #deselect first 3 rows (ID,LAT,LON)
    
    #remove NAs
    w<-!is.na(w_on)
    weights<-weights[w]
    weights<-weights/372
    w_on <- w_on[w]
  
    #calculate row-wise weighted mean winter onset
    mean_w_on<-weighted.mean(w_on,weights)
    
    #calculate row-wise weighted variance in winter onset
    #formula variance for weighed sample, using frequency weighed unbiased estimator (~ bessel's correction)
    upper <- sum(weights*(w_on-mean_w_on)^2)
    lower <- sum(weights)-1
    if (lower >0){wsd <- sqrt(upper/lower)} else {wsd<-NA} #if the whole year is NA, sum(weights) is  ==> lower = -1 ==> wsd <-NA
    #in that case mean_w_on becomes NAN, but that causes no trouble. upper simply becomes 0
    
    #store mean_w_on and wsd in dataframe

    weighted[[i]][1:length(input[[1]][,1]),1]<-input[[1]][,1] #assigns names
    weighted[[i]][j,2]<-mean_w_on
    weighted[[i]][j,3]<-wsd
  }
}


```



####plotting new weighted sd's
plotting the results: mean, sd and weighted sd (all weighted). sd has to be capped at 50, CV at 0.5 due to a long right-tail

```{r plot_cv}
input<-NA
for (i in 1:length(weighted)){
  subtitle<-paste(x[i],"days below ", y[i], "°C") #will be needed in figures
 # hist(weighted[[i]]$sd,breaks=100, main = paste("histogram standard deviation run ",i))
  input<-weighted[[i]]
  input$sd[input$sd>50]<-50
  input<-input[!is.na(input$m),]
  a<-merge(input,locations,by=1)
  a<-a[!is.na(a$sd),]
  plot(a$lat~a$lon,bg=rgb(0,a$m,0,maxColorValue=372),col=NA,main="weighted mean day of winter   onset",sub=subtitle,pch=22,cex=0.8,xlab="",ylab="") #subtitle is recycled from earlier chunk (unweighted calcs)
  plot(a$lat~a$lon,bg=rgb(0,a$sd,0,maxColorValue=max(a$sd)),col=NA,main="Winter variability, weighed",sub= subtitle,pch=22,cex=0.8,xlab="",ylab="")
  points(x=(-170):(-121),y= rep(25,50), col = rgb (0, 0:50,0,maxColorValue = 50),pch=22,cex=0.8)
  text(-170,22,"sd=0")
  text(-120,22,"sd = 50")
  a$CV<-a$sd/a$m
 # hist(a$CV,breaks=100,main =paste ("histogram of CV run ",i))
  a$CV[a$CV>0.5]<-0.5
  plot(a$lat~a$lon,bg=rgb(0,a$CV,0,maxColorValue=max(a$CV)),col=NA,main="weighted mean/weighted sd of winter onset",sub=subtitle,pch=22,cex=0.8,xlab="",ylab="")
}
```

sd or cv, which one to choose?


```{r}
for (i in 1:4){
plot(weighted[[i]]$sd ~ weighted[[i]]$m)
}
```
Because standard deviation does not increase with the mean, there is no need to use coefficient of variation (sd/m)


####some extremely preliminary models:

```{r models}
a$alt[a$alt==-999.9]<-NA
a$sd[a$sd>100]<-100
a<-a[!is.na(a$alt),]

do_models<-function(variable,name){
M<-lm(variable~a$lat+a$lon+a$alt)

M2<-lm(variable~a$lat)
plot(variable~a$lat,main = paste(name," vs lat"))#,sub=paste("slope ~ ",round(coef(M2)[2]*10,4)," per 10°"))
abline(M2,col=2,lwd=2)

M3<-lm(variable~a$lon)
plot(variable~a$lon,main = paste(name," vs lon"))#,sub=paste("slope ~ ",round(coef(M3)[2]*100,4)," per 100°"))
abline(M3,col=2,lwd=2)

M4<-lm(variable~a$alt)
plot(variable~a$alt,main = paste(name, " vs altitude"))#,sub=paste("slope ~ ",round(coef(M4)[2]*1000,4), "per 1000m"))
abline(M4,col=2,lwd=2)
}

do_models(a$m,"mean")
do_models(a$sd,"sd")
#do_models(a$CV, "cv")
```


The mean winter onset decreases (becomes earlier) at higher latitudes and decreases with altitude (earlier high up). SD decreases very slighlty with latitude and altitude.

One would expect that SD is only high when amplitude (difference winter-summer) is low ,because then a small drop in temperature counts as winter event. Based on amplitude patterns(pink graphics, way up), one expects SD to increase towards equator, and to increase towards oceans. Instead, high SDs appear mostly in central US. Why?


```{r}

me<-merge(a,reg,by.x=1,by.y=6) #lat.y and lon.y are wrong
me$A[me$A>300]<-300 # a few really extreme points
me$alt[me$alt==-999.9]<-NA
me<-me[!is.na(me$alt),]
mod<-lm(me$sd~me$A)
plot(me$sd~me$A,pch=22,cex=0.3)
abline(mod,col=2,lwd=2)


me<-me[me$sd>1,]
Ml<-lm(log(me$sd)~me$A)
plot(log(me$sd)~me$A,pch=22,cex=0.5)
abline(Ml,col=2,lwd=2)

#So the variance occurs indeed only in low-amplitude regions (though the green graphics suggested otherwise)

me$logsd<-log(me$sd)

plot(me$logsd~me$lat.x)
abline(lm(me$logsd~me$lat.x),col=2,lwd=2)
#that is strange, Amplitude correlates with latitude, so why does sd not correlate the same way 


sub<-me[me$lon.x<0,]
hi<-sub[sub$sd>quantile(me$sd,0.98),]
plot(sub$lat.x~sub$lon.x,bg=rgb(sub$A,0,sub$A,maxColorValue = max(sub$A)),pch=22,col=NA,cex=1.5,xlim=c(-130,-50))
points(hi$lat.x~hi$lon.x,cex=1.5,col=4,lwd=2)
##aah!}

```

The amplitude does not correlate strongly with latitude between 20°N and 60°N, the continentality is more important. high sd (like, highest 2% of dataset) occur rarely at really high amplitude regions (>50°N and central US), but mostly at the west coast (which has low amplitude) and the mountain ranges east to it (medium amplitude)


```{r extract_n}
ns<-list(NA,NA,NA,NA)
for (i in 1:length(lis)){
  ns[[i]]<-data.frame(NA,NA)
  for (rows in 1:nrow(lis[[i]][[1]])){
    ns[[i]][rows,1]<-sum(!is.na(lis[[i]][[1]][rows,4:147]))
    ns[[i]][rows,2]<-lis[[i]][[1]][rows,1]
  }
  names(ns[[i]])<-c("n","ID")
}
s<-merge(ns[[1]],me,by.x=2,by.y=1)
plot(s$n,s$sd)
sub<-s[s$lon.x<0,]
hi<-sub[sub$sd>quantile(me$sd,0.98),]
hi2<-sub[sub$n<4,]

plot(sub$lat.x~sub$lon.x,bg=rgb(0,0,sub$n,maxColorValue = max(sub$n)),pch=22,col=NA,cex=0.8)

#plot(hi$lon.x,hi$lat.x,bg=rgb(0,0,0),pch=22,col=NA,cex=0.8,main = "location unusally high variation")
points(hi2$lat.x~hi2$lon.x,cex=1.5,col=2,lwd=2)

plot(s$lat.x~s$lon.x,bg=1,pch=22,col=NA,cex=0.8)
n3<-s[s$n==3,]
points(n3$lat.x~n3$lon.x,col=2)
```



